% The document class supplies options to control rendering of some standard
% features in the result.  The goal is for uniform style, so some attention 
% to detail is *vital* with all fields.  Each field (i.e., text inside the
% curly braces below, so the MEng text inside {MEng} for instance) should 
% take into account the following:
%
% - author name       should be formatted as "FirstName LastName"
%   (not "Initial LastName" for example),
% - supervisor name   should be formatted as "Title FirstName LastName"
%   (where Title is "Dr." or "Prof." for example),
% - degree programme  should be "BSc", "MEng", "MSci", "MSc" or "PhD",
% - dissertation title should be correctly capitalised (plus you can have
%   an optional sub-title if appropriate, or leave this field blank),
% - dissertation type should be formatted as one of the following:
%   * for the MEng degree programme either "enterprise" or "research" to
%     reflect the stream,
%   * for the MSc  degree programme "$X/Y/Z$" for a project deemed to be
%     X%, Y% and Z% of type I, II and III.
% - year              should be formatted as a 4-digit year of submission
%   (so 2014 rather than the academic year, say 2013/14 say).

\documentclass[ % the name of the author
                    author={Lucas O'Dowd-Jones},
                % the name of the supervisor
                supervisor={Dr. Alex Kavvos},
                % the degree programme
                    degree={MEng},
                % the dissertation    title (which cannot be blank)
                     title={Variations on Normalisation by Evaluation in Haskell},
                % the dissertation subtitle (which can    be blank)
                  subtitle={},
                % the dissertation     type
                      type={programming languages},
                % the year of submission
                      year={2021}]{dissertation}

\usepackage{amsmath}
\usepackage{biblatex}
\usepackage{stmaryrd}
\addbibresource{bibliography.bib}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=[0]\bfseries\color{bittersweet},
  keywordstyle=[1]\bfseries\color{teal},
  keywordstyle=[3]\bfseries\color{dark},
  keywordstyle=[2]\bfseries,
  columns=spaceflexible,
  morekeywords=[0]{data, where, type, class, instance, case, of, do, return, infixr},
  morekeywords=[1]{Name, String, Expr, NormalForm, NeutralForm, Env, Map, V, NeutralV, FreshName, State, Set, DbExpr, Int, Ty, Nat, Vec, Elem, OPE, NeutralExpr, NormalExpr, forall, SingTy, STy, Any, SingContext, 'BaseTy, Eq, ChurchNumeralTy, ChurchNumeral, NFData, [Name], [Ty]},
  morekeywords=[2]{Z, S, True, case, of},
  morekeywords=[3]{ctx, ty, arg, result, ctx1, ctx2, strong, weak, ctxV, ctx', ctxV', strongest, a, b, c, arg1, arg2},
  mathescape=true,
  keepspaces=true,
  aboveskip=10pt,
  belowskip=10pt,
  alsoletter={', [, ]}
}

\newcommand{\code}{\lstinline}

\begin{document}
\definecolor{teal}{rgb}{0.0, 0.5, 0.5}
\definecolor{bittersweet}{rgb}{1.0, 0.44, 0.37}
\definecolor{dark}{rgb}{0.1, 0.45, 1.0}

% =============================================================================

% This macro creates the standard UoB title page by using information drawn
% from the document class (meaning it is vital you select the correct degree 
% title and so on).

\maketitle

% After the title page (which is a special case in that it is not numbered)
% comes the front matter or preliminaries; this macro signals the start of
% such content, meaning the pages are numbered with Roman numerals.

\frontmatter

% This macro creates the standard UoB declaration; on the printed hard-copy,
% this must be physically signed by the author in the space indicated.

\makedecl

% LaTeX automatically generates a table of contents, plus associated lists 
% of figures, tables and algorithms.  The former is a compulsory part of the
% dissertation, but if you do not require the latter they can be suppressed
% by simply commenting out the associated macro.

\tableofcontents

% \listoffigures
% \listoftables
% \listofalgorithms
% \lstlistoflistings

% The following sections are part of the front matter, but are not generated
% automatically by LaTeX; the use of \chapter* means they are not numbered.

% -----------------------------------------------------------------------------

\chapter*{Executive Summary}

% {\bf A compulsory section, of at most $1$ page} 

\noindent
This section should pr\'{e}cis the project context, aims and objectives,
and main contributions (e.g., deliverables) and achievements; the same 
section may be called an abstract elsewhere.  The goal is to ensure the 
reader is clear about what the topic is, what you have done within this 
topic, {\em and} what your view of the outcome is.

The former aspects should be guided by your specification: essentially 
this section is a (very) short version of what is typically the first 
chapter.  Note that for research-type projects, this {\bf must} include 
a clear research hypothesis.  This will obviously differ significantly
for each project, but an example might be as follows:

\begin{quote}
My research hypothesis is that a suitable genetic algorithm will yield
more accurate results (when applied to the standard ACME data set) than 
the algorithm proposed by Jones and Smith, while also executing in less
time.
\end{quote}

\noindent
The latter aspects should (ideally) be presented as a concise, factual 
bullet point list.  Again the points will differ for each project, but 
an might be as follows:

\begin{quote}
\noindent
\begin{itemize}
\item I spent $120$ hours collecting material on and learning about the 
      Java garbage-collection sub-system. 
\item I wrote a total of $5000$ lines of source code, comprising a Linux 
      device driver for a robot (in C) and a GUI (in Java) that is 
      used to control it.
\item I designed a new algorithm for computing the non-linear mapping 
      from A-space to B-space using a genetic algorithm, see page $17$.
\item I implemented a version of the algorithm proposed by Jones and 
      Smith in [6], see page $12$, corrected a mistake in it, and 
      compared the results with several alternatives.
\end{itemize}
\end{quote}

% -----------------------------------------------------------------------------

\chapter*{Acknowledgements}

% {\bf An optional section, of at most $1$ page}

\noindent
It is common practice (although totally optional) to acknowledge any
third-party advice, contribution or influence you have found useful
during your work.  Examples include support from friends or family, 
the input of your Supervisor and/or Advisor, external organisations 
or persons who  have supplied resources of some kind (e.g., funding, 
advice or time), and so on.

% =============================================================================

% After the front matter comes a number of chapters; under each chapter,
% sections, subsections and even subsubsections are permissible.  The
% pages in this part are numbered with Arabic numerals.  Note that:
%
% - A reference point can be marked using \label{XXX}, and then later
%   referred to via \ref{XXX}; for example Chapter\ref{chap:context}.
% - The chapters are presented here in one file; this can become hard
%   to manage.  An alternative is to save the content in seprate files
%   the use \input{XXX} to import it, which acts like the #include
%   directive in C.

\mainmatter

% -----------------------------------------------------------------------------

\input{chapters/Introduction.tex}

% -----------------------------------------------------------------------------

\input{chapters/UntypedNbE.tex}

% -----------------------------------------------------------------------------

\input{chapters/TypedLambdaCalculus.tex}

% -----------------------------------------------------------------------------

\input{chapters/TypedNbE.tex}

% -----------------------------------------------------------------------------

\input{chapters/CriticalEvaluation.tex}

% -----------------------------------------------------------------------------

\input{chapters/Conclusion.tex}

\backmatter

\printbibliography

% -----------------------------------------------------------------------------

% The dissertation concludes with a set of (optional) appendicies; these are 
% the same as chapters in a sense, but once signaled as being appendicies via
% the associated macro, LaTeX manages them appropriatly.

\appendix

\chapter{An Example Appendix}
\label{appx:example}

Content which is not central to, but may enhance the dissertation can be 
included in one or more appendicies.

\noindent
Note that in line with most research conferences, the marking panel is not
obliged to read such appendices.

% =============================================================================

\end{document}
